Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2016-08-31 LaTeX/Turnover.tex
lambdas added

%life-cycle
The life-cycle begins with haploid gametes/gametophytes experiencing selection that depends on their allele at the \textbf{A} locus.
Gametes/gametophytes from males then mate randomly with gametes/gametophytes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci.
There is then a period of diploid selection that again depends on genotypes at the \textbf{A} locus, followed by meiosis with recombination. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

@@

%resident equilibrium
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p_{Xf}$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p_{Xm}$, and Y-bearing, $p_{Ym}$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
Without haploid selection this reduces to the results of \citep{vanDoorn:2010hu}.

%external stability/turnover

>>>>

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes within each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes/gametophytes from males then mate randomly with gametes/gametophytes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci.
Diploids then compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes/gametophytes from the homogametic sex (e.g., in eggs from an XX female, $p_{Xf}$) and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p_{Xm}$, and Y-bearing, $p_{Ym}$).
Assuming selection is weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete/gametophyte are small, and we can solve for the mean frequency of $A$ across all types ($p_A$) as well as the difference in the frequencies of $A$ between two of the three types, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
With no haploid selection and $k=0$ (neo-$Y$ invading a $ZW$ system) or $k=1$ (neo-$W$ invading an $XY$ system), our results reduce to those of \cite{vanDoorn:2010hu}.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the results for invasion into an $XY$ system assuming equal dominance coefficients in the two sexes, no competition among gametes/gametophytes from females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}.
Equivalent results for invasions into a $ZW$ system are given in the Sup.\ Mat.

\subsubsection*{Neo-$Y$}

The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
\begin{equation}
\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
\end{equation}
where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%write in general form without weak selection

\subsubsection*{Neo-$W$}

The growth rate of a rare, dominant neo-$W$ region ($k=1$) is
\begin{equation}
\lambda_{W,XY} = 1 + V_A \frac{s_f \left( (2 r (1-R) +R) s_f + (1- 2r) R s_m \right)}{2 r R \left( s_f + s_m \right)^2} t_m^2.
\end{equation}
In this case, even when the novel sex-determining locus is in looser linkage with the selected locus than is the ancestral sex-determining locus, $r<R$, a novel sex-determiner can invade.
For example, with $R=1/2$ the neo-W invades if there is some linkage between the ancestral sex-determining and selected loci ($r<1/2$), there is selection among gametes/gametophytes in males ($t_m\neq0$), and there is selection for or against $A$ in both males and females ($s_m s_f >0$) that is stronger in females than in males ($|s_f| > |s_m|$).
%write in general form without weak selection

\subsubsection*{Neo-ESD}

The growth rate of a rare, dominant neo-ESD region that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD,XY} = 1 + \frac{(\lambda_{Y,XY} - 1) + (\lambda_{W,XY} -1)}{4} \Big|_{R=1/2}
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s_m - 3 s_f\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\end{equation}
Thus the $k=1/2$ case gets half of the advantages of a neo-$W$ and half that of a neo-$Y$.  
Recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
%write in general form without weak selection
